% -*- Mode: LaTeX -*-

% overview.tex --


% Copyright (c) 1996, 1997 Regents of the University of California.
% All rights reserved.
% 
% Redistribution and use in source and binary forms, with or without
% modification, are permitted provided that the following conditions
% are met:
% 1. Redistributions of source code must retain the above copyright
%    notice, this list of conditions and the following disclaimer.
% 2. Redistributions in binary form must reproduce the above copyright
%    notice, this list of conditions and the following disclaimer in the
%    documentation and/or other materials provided with the distribution.
% 3. All advertising materials mentioning features or use of this software
%    must display the following acknowledgement:
%      This product includes software developed by the University of
%      California, Berkeley and the California PATH Program.
% 4. Neither the name of the University nor of the California PATH
%    Program may be used to endorse or promote products derived from
%    this software without specific prior written permission.
% 
% THIS SOFTWARE IS PROVIDED BY THE REGENTS AND CONTRIBUTORS ``AS IS'' AND
% ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
% IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE
% ARE DISCLAIMED.  IN NO EVENT SHALL THE REGENTS OR CONTRIBUTORS BE LIABLE
% FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL
% DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS
% OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION)
% HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT
% LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY
% OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF
% SUCH DAMAGE.

\section{Overview\label{overview}} \todaysname{} is a description
language for dynamic
networks of hybrid systems.  Such systems consist of components which
can be created, interconnected and destroyed as the system evolves in
time.  Components exhibit hybrid behavior, consisting of
continuous-time phases separated by discrete-event transitions.  They
may evolve independently or jointly with each other.  They interact
through input-output connections and synchronization of events.  The
interaction network may evolve in time.

We believe this model offers the proper level of abstraction for
describing complex applications such as highway and air traffic
control systems, robotic shopfloors, and other systems whose operation
cannot be captured by conventional models, which typically support
only static interconnections.

\todaysname{} is extensible.  We plan to use \todaysname{} as a
mechanism for defining the Automated Highway System Tool Interchange
Format (AHSTIF).  An automated highway system is a hybrid system with
specific characteristics and restrictions. \todaysname{} supports the
definition and enforcement of AHSTIF-specific semantics.

This is a reference manual for \todaysname.  It presents its material
in a logical order, but it aims to be concise and has no pretense to
be a \todaysname{} tutorial.  Section~\ref{model} describes the model,
and section~\ref{language} the language.

The name \todaysname{} is a permutation of HSTIF (Hybrid System Tool
Interchange Format).

